---
title: "Figure 4 – NDD Gene × Perturbation Analysis"
author: ""
date: "`r Sys.Date()`"
output:
  html_notebook:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 5,
  dpi        = 150
)
```

## Configuration

**Edit `DATA_DIR` to match your environment before running.**

- `DATA_DIR` → `/data/wholebrain_crispr_atlas/analysis/` folder
- To update:
  - `celltype_pert_number_of_cells.csv` (required for Fig 4B row annotation)

```{r config}
# ── Path to the bundled input files ───────────────────────────────────────────
DATA_DIR <- "/data/wholebrain_crispr_atlas/analysis"

# ── Output directory for saved PDFs ───────────────────────────────────────────
OUT_DIR <- "./Figure4_output"
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)

stopifnot(dir.exists(DATA_DIR))
message("DATA_DIR: ", DATA_DIR)
```

## Libraries

```{r libraries}
suppressPackageStartupMessages({
  library(tidyverse)   # dplyr, ggplot2, tidyr, forcats, readr, purrr
  library(data.table)
  library(readxl)
  library(arrow)       # read_parquet
  library(patchwork)
  library(cowplot)
  library(ggbeeswarm)
  library(ggpubr)
  library(ggrepel)
  library(pheatmap)
  library(viridis)
  library(scales)
  library(gtable)
  library(grid)
})
```

## Shared data loading

These objects are reused across multiple figures.

```{r shared-data}
# ── Gene database ──────────────────────────────────────────────
annot_path <- file.path(DATA_DIR, "ndd_perturbation", "geneDB-raw-data_Feb2026_Epi2_ndd_rbp_PA_xin_only_main_select.xlsx")
stopifnot(file.exists(annot_path))

annot_raw <- read_excel(annot_path)


# Sanitize column names (replace "-" with ".") for safe $ access
annot <- annot_raw
colnames(annot) <- make.names(colnames(annot))

# ── DEG results ───────────────────────────────────────────────────────────────
deg_path <- file.path(DATA_DIR, "diff_expr_genes", "wilcoxon_de_results_group_name_final_data_no_multi_guide_cells.parquet")
stopifnot(file.exists(deg_path))

de_res <- read_parquet(deg_path) %>% filter(pvals_adj < 0.05)
de_res <- de_res[!grepl("Safe_target", de_res$gene_target), ]

message(sprintf("Loaded %d DEG rows across %d perturbation targets",
                nrow(de_res), length(unique(de_res$gene_target))))

# ── Shared NDD gene list (using sanitized column names) ───────────────────────
build_ndd_list <- function(sl) {
  list(
    ASD185          = unique(sl$Mouse.gene[sl$asc.denovo.asd185 == TRUE]),
    NDD664          = unique(sl$Mouse.gene[sl$asc.denovo.ndd664 == TRUE]),
    DD285           = unique(sl$Mouse.gene[sl$deciphering.developmental.disorders.ddd.significant == TRUE]),
    GEL_ID          = unique(sl$Mouse.gene[sl$gel_id == 10]),
    GEL_EPI         = unique(sl$Mouse.gene[sl$gel_epi == 10]),
    GEL_ATAX        = unique(sl$Mouse.gene[sl$gel_atax == 5]),
    SCZ244          = unique(sl$Mouse.gene[sl$asc.denovo.scz244 == TRUE]),
    ADHD            = unique(sl$Mouse.gene[sl$adhd.demontis.2021.count > 0]),
    BPD             = unique(sl$Mouse.gene[sl$bdp.mullins.2021.count > 0]),
    MDD             = unique(sl$Mouse.gene[sl$mdd.levey.2021.count > 0]),
    AD              = unique(sl$Mouse.gene[sl$ad.bellenguez.2022.count > 0]),
    PD              = unique(sl$Mouse.gene[sl$Parkinsons_Evidence != "."]),
    ALS             = unique(sl$Mouse.gene[sl$ALS_Evidence != "."]),
    LoF_tolerant    = unique(sl$Mouse.gene[sl$gnomadv2.lof.oe.ratio.upper.decile.bin %in% c("6","7","8","9")]),
    LoF_intermediate = unique(sl$Mouse.gene[sl$gnomadv2.lof.oe.ratio.upper.decile.bin %in% c("3","4","5")]),
    LoF_constrained = unique(sl$Mouse.gene[sl$gnomadv2.lof.oe.ratio.upper.decile.bin %in% c("0","1","2")]),
    Blomen_HAP1     = unique(sl$Mouse.gene[sl$gene.essentiality.blomen.2015.hap1.selected == TRUE]),
    Blomen_KBM7     = unique(sl$Mouse.gene[sl$gene.essentiality.blomen.2015.kbm7.selected == TRUE])
  )
}

ref_ndd_list <- build_ndd_list(annot)

# ── Shared disorder class labels and colours ──────────────────────────────────
DISORDER_LABELS <- c(
  ASD185 = "ASD (n=185, Fu 2022)", NDD664 = "NDD (n=664, Fu 2022)",
  DD285  = "DD (n=285, Kaplanis 2020)", GEL_ID = "GEL ID (Green)",
  GEL_EPI = "GEL Epilepsy (Green)", GEL_ATAX = "GEL Ataxia (Green)",
  SCZ244 = "SCZ (n=244, Singh 2022)", ADHD = "ADHD (Demontis 2021)",
  BPD = "Bipolar (Mullins 2021)", MDD = "MDD (Levey 2021)",
  AD = "AD (Bellenguez 2022)", PD = "GEL PD (Green)", ALS = "GEL ALS (Green)",
  LoF_tolerant = "LoF tolerant", LoF_intermediate = "LoF intermediate",
  LoF_constrained = "LoF constrained",
  Blomen_HAP1 = "Blomen HAP1", Blomen_KBM7 = "Blomen KBM7"
)

DISORDER_CLASS <- c(
  ASD185="Neurodevelopmental", NDD664="Neurodevelopmental", DD285="Neurodevelopmental",
  GEL_ID="Neurodevelopmental", GEL_EPI="Neurodevelopmental", GEL_ATAX="Neurodevelopmental",
  SCZ244="Neuropsychiatric", ADHD="Neuropsychiatric", BPD="Neuropsychiatric", MDD="Neuropsychiatric",
  AD="Neurodegenerative", PD="Neurodegenerative", ALS="Neurodegenerative",
  LoF_tolerant="LOEUF", LoF_intermediate="LOEUF", LoF_constrained="LOEUF",
  Blomen_HAP1="Essential", Blomen_KBM7="Essential"
)

CLASS_COLOURS <- c(
  Neurodevelopmental = "#FB9A99",
  Neuropsychiatric   = "#8DD3C7",
  Neurodegenerative  = "#BEBADA",
  LOEUF              = "#91d288",
  Essential          = "#86b3db"
)

DISORDER_ORDER <- c(
  "ASD185","NDD664","DD285","GEL_ID","GEL_EPI","GEL_ATAX",
  "SCZ244","BPD","MDD","ADHD",
  "AD","PD","ALS",
  "LoF_tolerant","LoF_intermediate","LoF_constrained",
  "Blomen_HAP1","Blomen_KBM7"
)

CELLTYPE_ORDER <- c(
  "007 L2-3 IT CTX Glut", "005 L4-5 IT CTX Glut",
  "008 L2-3 IT ENT PPP RSP Glut", "009 L2-3 IT PIR AON ENT Glut",
  "022 L5 ET CTX Glut", "001 L5-6 IT Glut", "027 NP-CT-L6b-OB Glut",
  "046 CTX-CGE GABA", "052 Pvalb Gaba", "053 Sst Gaba",
  "017 CA3 CA2-FC DG Glut", "012 MEA LA CA1 DG Glut",
  "151 TH Prkcd Grin2c Glut", "145 MH-LH TH Glut",
  "110 CNU-HYa HY MM Glut", "066 CNU-HYa HY GABA",
  "155 MB Glut", "215 MB Dopa", "191 MB P MY GABA",
  "054 CNU-MGE GABA", "059 CNU-LGE LSX GABA",
  "308 CB GABA", "217 P MY Pineal Glut"
)
```

---

## Figure 4A – Risk gene counts by disorder category

```{r fig4a, fig.width=7, fig.height=6}
summary_counts <- tibble(
  short_name    = names(ref_ndd_list),
  category      = DISORDER_LABELS[names(ref_ndd_list)],
  disorder_class = DISORDER_CLASS[names(ref_ndd_list)],
  n_genes       = sapply(ref_ndd_list, length)
)

fig4a_df <- summary_counts %>%
  mutate(
    disorder_class = factor(disorder_class,
                            levels = c("Neurodevelopmental","Neuropsychiatric",
                                       "Neurodegenerative","LOEUF","Essential")),
    category = fct_reorder(category, n_genes)
  )

fig4a <- ggplot(fig4a_df, aes(x = category, y = n_genes, fill = disorder_class)) +
  geom_col(width = 0.8, color = "black", linewidth = 0.2) +
  geom_text(aes(label = n_genes), hjust = -0.15, size = 3) +
  facet_grid(disorder_class ~ ., scales = "free_y", space = "free_y") +
  scale_fill_manual(values = CLASS_COLOURS) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  coord_flip() +
  theme_classic(base_size = 10) +
  xlab("") + ylab("Number of risk genes") +
  theme(
    axis.text.y  = element_text(size = 8),
    legend.position = "none",
    strip.text   = element_text(face = "bold", size = 10)
  )

fig4a

# ggsave(fig4a, filename = file.path(OUT_DIR, "Fig4A_Count_Disorder_Riskgenes.pdf"),
#       width = 6, height = 6)
```

---

## Figure 4B – DEG burden heatmap (cell type × disorder)

> **Note:** The row annotation (total cells per cell type) requires
> `celltype_pert_number_of_cells.csv`. If the file is absent the heatmap
> renders without that annotation bar.

```{r fig4b, fig.width=10, fig.height=7}
# Count DEGs per cell type × perturbation
deg_counts <- de_res %>%
  group_by(group_name, gene_target) %>%
  summarise(n_degs = n(), .groups = "drop")

# Map perturbations to disorder categories (long-form join)
ref_map <- stack(ref_ndd_list) %>%
  rename(gene_target = values, disorder = ind)

deg_disorder <- deg_counts %>%
  inner_join(ref_map, by = "gene_target") %>%
  group_by(group_name, disorder) %>%
  summarise(total_degs = sum(n_degs), .groups = "drop")

# Pivot to matrix; log10 transform
mat_df <- deg_disorder %>%
  pivot_wider(names_from = disorder, values_from = total_degs, values_fill = 0)

mat <- as.matrix(mat_df[, -1])
rownames(mat) <- mat_df$group_name
mat_log <- log10(mat + 1)

# Cap at 95th percentile to reduce outlier
cap <- quantile(mat_log, 0.95)
mat_log[mat_log > cap] <- cap

# Reorder rows/cols to match desired biological order
avail_ct       <- intersect(CELLTYPE_ORDER, rownames(mat_log))
avail_disorder <- intersect(DISORDER_ORDER, colnames(mat_log))
mat_log        <- mat_log[avail_ct, avail_disorder, drop = FALSE]

# Disorder category annotation (rows of transposed matrix = disorders)
disorder_anno <- data.frame(
  Disorder_Type = DISORDER_CLASS[avail_disorder],
  row.names     = avail_disorder
)

disorder_anno_colors <- list(
  Disorder_Type = CLASS_COLOURS
)

# Optional cell-count annotation
cell_count_path <- file.path(DATA_DIR, "celltype_pert_number_of_cells.csv")
col_anno_arg <- NULL
if (file.exists(cell_count_path)) {
  cell_counts <- read.csv(cell_count_path,
                          col.names = c("cell_type","perturbation",
                                        "cell_type_x_pert","number_of_cells"))
  total_cells <- cell_counts %>%
    group_by(cell_type) %>%
    summarise(total_cells = sum(number_of_cells, na.rm = TRUE), .groups = "drop") %>%
    filter(cell_type %in% avail_ct) %>%
    column_to_rownames("cell_type")
  col_anno_arg <- total_cells[avail_ct, , drop = FALSE]
} else {
  message("celltype_pert_number_of_cells.csv not found; skipping cell-count annotation.")
}

fig4b <- pheatmap(
  t(mat_log),
  color          = viridis(100),
  cluster_rows   = FALSE,
  cluster_cols   = FALSE,
  show_rownames  = TRUE,
  show_colnames  = TRUE,
  fontsize_row   = 9,
  fontsize_col   = 8,
  border_color   = NA,
  angle_col      = 315,
  main           = "Cell type-specific perturbation effects (log10 DEGs)",
  annotation_row = disorder_anno,
  annotation_col = col_anno_arg,
  annotation_colors = disorder_anno_colors,
  gaps_row       = c(6, 10, 13, 16),
  gaps_col       = c(7, 10, 12, 14, 16, 19, 21, 22),
  silent         = TRUE
)

grid.newpage()
grid.draw(fig4b$gtable)

# pdf(file.path(OUT_DIR, "Fig4B_DEGBurden_Heatmap.pdf"), width = 10, height = 7)
# grid.newpage(); grid.draw(fig4b$gtable)
# dev.off()
```

---

## Figure 4C – DEG burden distribution (ASD/DD + LoF constraint + Essentiality)

```{r fig4c, fig.width=9, fig.height=4.5}
# Merge DEG results with gene database
de_merged <- merge(de_res, annot, by.x = "gene_target", by.y = "Mouse.gene",
                   all.x = TRUE)

gene_level <- de_merged %>%
  group_by(gene_target) %>%
  mutate(total_DEG_count = n()) %>%
  ungroup() %>%
  distinct(gene_target, total_DEG_count, .keep_all = TRUE) %>%
  mutate(
    log_count = log10(total_DEG_count),
    ASD185 = ifelse(asc.denovo.asd185 == TRUE, "Case", "Non-case"),
    DD285  = ifelse(deciphering.developmental.disorders.ddd.significant == TRUE,
                   "Case", "Non-case"),
    LoF_group = case_when(
      gnomadv2.lof.oe.ratio.upper.decile.bin %in% c("6","7","8","9") ~ "LoF_tolerant",
      gnomadv2.lof.oe.ratio.upper.decile.bin %in% c("3","4","5")     ~ "LoF_intermediate",
      gnomadv2.lof.oe.ratio.upper.decile.bin %in% c("0","1","2")     ~ "LoF_constrained"
    ),
    LoF_group = factor(LoF_group,
                       levels = c("LoF_tolerant","LoF_intermediate","LoF_constrained"))
  )

# ── Panel: ASD185 / DD285 ────────────────────────────────────────────────────
neurodev_long_fixed <- gene_level %>%
  dplyr::select(gene_target, log_count, total_DEG_count, ASD185, DD285) %>%
  filter(ASD185 == "Case" | DD285 == "Case") %>%
  mutate(in_ASD185 = !is.na(ASD185) & ASD185 == "Case",
         in_DD285  = !is.na(DD285)  & DD285  == "Case") %>%
  pivot_longer(cols = c(in_ASD185, in_DD285),
               names_to = "disorder", values_to = "is_case") %>%
  filter(is_case) %>%
  mutate(disorder = case_when(disorder == "in_ASD185" ~ "ASD185",
                              disorder == "in_DD285"  ~ "DD285"))

mean_fixed <- neurodev_long_fixed %>%
  group_by(disorder) %>%
  summarise(mean_log = mean(log_count, na.rm = TRUE),
            mean_raw = mean(total_DEG_count, na.rm = TRUE),
            n = n(), .groups = "drop")

p_neurodev_fixed <- ggplot(neurodev_long_fixed, aes(x = log_count)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "#FB9A99", alpha = 0.8, color = "black") +
  geom_density(color = "#FB9A99", fill = "#FB9A99", linewidth = 1, alpha = 0.4) +
  geom_vline(data = mean_fixed, aes(xintercept = mean_log), linetype = "dashed") +
  geom_text(data = mean_fixed,
            aes(x = mean_log, y = Inf,
                label = paste0("Mean = ", round(mean_raw, 1), "  (n=", n, ")")),
            vjust = 1.5, hjust = -0.1, inherit.aes = FALSE, size = 3) +
  facet_wrap(~disorder, ncol = 1, scales = "free_y") +
  theme_classic(base_size = 10) +
  labs(x = "log10(#Total DEGs)", y = NULL) +
  ggtitle("Neurodevelopmental (pivot_longer fix)")

p_neurodev_fixed
```
---


```{r fig4c-rest, fig.width=9, fig.height=4.5}
# ── Helper: distribution-split histogram ─────────────────────────────────────
plot_split_hist <- function(df, group_col, fill_map = NULL, title = NULL) {
  mean_df <- df %>%
    group_by(.data[[group_col]]) %>%
    summarise(mean_log = mean(log_count, na.rm = TRUE),
              mean_raw = mean(total_DEG_count, na.rm = TRUE), .groups = "drop")

  p <- ggplot(df, aes(x = log_count,
                      fill  = .data[[group_col]],
                      color = .data[[group_col]])) +
    geom_histogram(aes(y = after_stat(density)),
                   alpha = 0.6, bins = 30, position = "identity",
                   color = "black") +
    geom_density(alpha = 0.3, linewidth = 1) +
    geom_vline(data = mean_df, aes(xintercept = mean_log), linetype = "dashed") +
    geom_text(data = mean_df,
              aes(x = mean_log, y = Inf,
                  label = paste0("Mean = ", round(mean_raw, 1))),
              vjust = 1.75, hjust = -0.1, size = 3, inherit.aes = FALSE) +
    facet_wrap(as.formula(paste("~", group_col)), ncol = 1, scales = "free_y") +
    theme_classic(base_size = 10) +
    labs(x = "log10(#Total DEGs)") +
    theme(legend.position = "none",
          strip.text = element_text(size = 10),
          axis.title.y = element_blank())
  if (!is.null(fill_map)) {
    p <- p + scale_fill_manual(values = fill_map) +
             scale_color_manual(values = fill_map)
  }
  if (!is.null(title)) p <- p + ggtitle(title)
  p
}

# ── Panel: LoF constraint ─────────────────────────────────────────────────────
p_lof <- plot_split_hist(
  df        = gene_level %>% filter(!is.na(LoF_group)),
  group_col = "LoF_group",
  fill_map  = c(LoF_tolerant = "#B2182B", LoF_intermediate = "grey50",
                LoF_constrained = "#2C7FB8"),
  title     = "LoF constraint"
)

# ── Panel: Blomen HAP1 essentiality ───────────────────────────────────────────
gene_level_blomen <- gene_level %>%
  filter(!is.na(gene.essentiality.blomen.2015.hap1.selected),
         gene.essentiality.blomen.2015.hap1.selected != ".") %>%
  mutate(gene.essentiality.blomen.2015.hap1.selected =
           as.character(gene.essentiality.blomen.2015.hap1.selected))

p_blomen <- plot_split_hist(
  df        = gene_level_blomen,
  group_col = "gene.essentiality.blomen.2015.hap1.selected",
  fill_map  = c("FALSE" = "grey50", "TRUE" = "#A1D99B"),
  title     = "Blomen 2015 HAP1"
)

fig4c <- plot_grid(
  p_neurodev_fixed,
  plot_grid(p_blomen, p_lof, nrow = 1, align = "hv"),
  nrow = 1, rel_widths = c(1, 2), align = "h"
)

# Statistical tests
cat("\n--- Kruskal-Wallis: LoF group ---\n")
print(kruskal.test(total_DEG_count ~ LoF_group,
                   data = gene_level %>% filter(!is.na(LoF_group))))

cat("\n--- Pairwise Wilcoxon (BH): LoF group ---\n")
print(pairwise.wilcox.test(
  gene_level$total_DEG_count[!is.na(gene_level$LoF_group)],
  gene_level$LoF_group[!is.na(gene_level$LoF_group)],
  p.adjust.method = "BH"
))

cat("\n--- Wilcoxon: Blomen HAP1 essential vs non-essential ---\n")
blomen_test_df <- gene_level %>%
  filter(!is.na(gene.essentiality.blomen.2015.hap1.selected),
         gene.essentiality.blomen.2015.hap1.selected != ".")
print(wilcox.test(total_DEG_count ~ gene.essentiality.blomen.2015.hap1.selected,
                  data = blomen_test_df))

# ggsave(fig4c, filename = file.path(OUT_DIR, "Fig4C_DEGBurden_Distribution.pdf"),
#        width = 9, height = 4.5)

fig4c
```

---

## Figure 4D – DEG burden vs inheritance (dominant / recessive / X-linked) and TADA FDR

```{r fig4d, fig.width=7, fig.height=5}
base_theme <- theme_classic(base_size = 8) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.line  = element_line(linewidth = 0.6),
    axis.ticks = element_line(linewidth = 0.6)
  )

score_palette <- c(Dominant = "#2C7FB8", Recessive = "#1B9E77", "X-linked" = "#6A3D9A")

# ── Helper: build group df for one inheritance score ─────────────────────────
make_score_df <- function(score_col, min_score, score_type) {
  de_merged %>%
    group_by(gene_target, .data[[score_col]]) %>%
    summarise(total_deg = n(), .groups = "drop") %>%
    mutate(log_total_deg = log10(total_deg)) %>%
    filter(.data[[score_col]] > min_score) %>%
    { med <- median(.[[score_col]], na.rm = TRUE)
      mutate(., group = ifelse(.data[[score_col]] > med, "Top", "Bottom"),
             score_type = score_type) }
}

dom_df <- make_score_df("dom_score", 20, "Dominant")
rec_df <- make_score_df("rec_score", 10, "Recessive")
x_df   <- de_merged %>%
  filter(gencode.chromosome == "chrX") %>%
  group_by(gene_target, xlink_score) %>%
  summarise(total_deg = n(), .groups = "drop") %>%
  mutate(log_total_deg = log10(total_deg)) %>%
  { med <- median(.$xlink_score, na.rm = TRUE)
    mutate(., group = ifelse(xlink_score > med, "Top", "Bottom"),
           score_type = "X-linked") }

combined_inh <- bind_rows(dom_df, rec_df, x_df) %>%
  mutate(
    group      = factor(group, levels = c("Top", "Bottom")),
    score_type = factor(score_type, levels = c("Dominant","Recessive","X-linked"))
  )

# p-values for Top vs Bottom within each score_type (no ggpubr)
p_vals_inh <- combined_inh %>%
  group_by(score_type) %>%
  summarise(p_val = t.test(log_total_deg ~ group)$p.value, .groups = "drop") %>%
  mutate(y = 4, label = sprintf("p = %.2g", p_val))

p_inh <- ggplot(combined_inh, aes(x = group, y = log_total_deg,
                                   fill = score_type, alpha = group)) +
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(aes(group = group), method = "t.test",
                     label = "p.format", size = 2, label.y = 4) +
  geom_jitter(width = 0.15, size = 0.5, alpha = 0.4, color = "black") +
  facet_wrap(~score_type) +
  scale_fill_manual(values = score_palette) +
  scale_alpha_manual(values = c(Top = 1, Bottom = 0.3)) +
  base_theme +
  labs(x = "", y = "log10(#Total DEGs)") +
  theme(legend.position = "none")


# ── TADA FDR panels ───────────────────────────────────────────────────────────
make_tada_df <- function(tada_col) {
  de_merged %>%
    group_by(gene_target, .data[[tada_col]]) %>%
    summarise(total_deg = n(), .groups = "drop") %>%
    mutate(log_total_deg = log10(total_deg) + 1) %>%
    filter(.data[[tada_col]] != ".") %>%
    mutate(
      fdr_val   = as.numeric(.data[[tada_col]]),
      logfdr    = -log10(fdr_val),
      logfdr    = ifelse(is.infinite(logfdr), 20, logfdr),
      logfdr_group = ntile(logfdr, 5),
      logfdr_group = factor(logfdr_group, levels = 1:5,
                            labels = c("Q1 (lowest)","Q2","Q3","Q4","Q5 (highest)"))
    )
}

make_tada_plot <- function(tada_df, fill_vals, x_label) {
  pw <- pairwise.t.test(tada_df$log_total_deg, tada_df$logfdr_group,
                        p.adjust.method = "BH")
  pw_df <- as.data.frame(as.table(pw$p.value)) %>%
    rename(group1 = Var1, group2 = Var2, p.adj = Freq) %>%
    filter(!is.na(p.adj), p.adj < 0.05) %>%
    mutate(
      label      = cut(p.adj, breaks = c(-Inf,.001,.01,.05,Inf),
                       labels = c("***","**","*","ns")),
      y.position = max(tada_df$log_total_deg, na.rm = TRUE) +
                     seq(0.2, by = 0.2, length.out = n())
    ) %>%
    filter(label != "ns")
  if (nrow(pw_df) > 0) pw_df$x <- (as.numeric(pw_df$group1) + as.numeric(pw_df$group2)) / 2

  ggplot(tada_df, aes(x = logfdr_group, y = log_total_deg, fill = logfdr_group)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.8, color = "black", linewidth = 0.4) +
    geom_jitter(width = 0.15, alpha = 0.25, size = 0.2, color = "black") +
    { if (nrow(pw_df) > 0) geom_text(data = pw_df, aes(x = x, y = y.position, label = label),
                                       inherit.aes = FALSE, size = 2.5) else list() } +
    scale_fill_manual(values = fill_vals) +
    theme_classic(base_size = 8) +
    labs(x = x_label, y = expression(log[10]("#Total DEGs"))) +
    theme(legend.position = "none")
}

asd_fills <- c("#FFF7BC","#80B1D3","#8DD3C7","#02818A","#08519C")
ndd_fills <- c("#FCCDE5","#FCBBA1","#FB9A99","#C44569","#BF812E")

p_tada_asd <- make_tada_plot(
  make_tada_df("asc.denovo.fdr.tada.asd"), asd_fills,
  expression(-log[10](FDR)~quintile~"(ASD)")
)
p_tada_ndd <- make_tada_plot(
  make_tada_df("asc.denovo.fdr.tada.ndd"), ndd_fills,
  expression(-log[10](FDR)~quintile~"(NDD)")
)

fig4d <- plot_grid(
  p_inh,
  plot_grid(p_tada_asd, p_tada_ndd, nrow = 1),
  ncol = 1
)

fig4d

# ggsave(fig4d, filename = file.path(OUT_DIR, "Fig4D_Inheritance_TADA.pdf"),
#        width = 7, height = 5)
```

---

## Figure 4F – DEG heatmap: 2k perturb × cell type (top DEG genes)

```{r fig4f, fig.width=11, fig.height=8}
# Number of significant DEGs (adj p < 0.1) in each cell type x perturbation combo
ndd_deg_path <- file.path(DATA_DIR, "ndd_perturbation", "celltype_NDD_gene_No_DEGs.csv")
stopifnot(file.exists(ndd_deg_path))

ndd_deg <- read.csv(ndd_deg_path)

# Aggregate duplicates
ndd_deg_clean <- ndd_deg %>%
  group_by(group_name, gene_target) %>%
  summarise(n_sig = sum(n_sig), .groups = "drop")

# Keep genes with ≥ 10 total DEGs across all cell types
gene_filter <- ndd_deg_clean %>%
  group_by(gene_target) %>%
  summarise(total_sig = sum(n_sig), .groups = "drop") %>%
  filter(total_sig >= 10)

ndd_deg_filt <- ndd_deg_clean %>%
  filter(gene_target %in% gene_filter$gene_target)

# Pivot to matrix and 0–1 normalise per gene (column)
mat_f_df <- ndd_deg_filt %>%
  pivot_wider(names_from = gene_target, values_from = n_sig, values_fill = 0) %>%
  as.data.frame()

rownames(mat_f_df) <- mat_f_df$group_name
mat_f_df$group_name <- NULL
mat_f <- as.matrix(mat_f_df)

col_range <- apply(mat_f, 2, function(x) { r <- diff(range(x)); if (r == 0) 1 else r })
mat_f_norm <- sweep(sweep(mat_f, 2, apply(mat_f, 2, min), "-"), 2, col_range, "/")

# Row annotation: total cells per cell type (optional)
cell_count_path <- file.path(DATA_DIR, "celltype_pert_number_of_cells.csv")
row_anno_f <- NULL
if (file.exists(cell_count_path)) {
  cell_counts_f <- read.csv(cell_count_path,
                             col.names = c("cell_type","perturbation",
                                           "cell_type_x_pert","number_of_cells"))
  total_cells_f <- cell_counts_f %>%
    group_by(cell_type) %>%
    summarise(total_cells = sum(number_of_cells, na.rm = TRUE), .groups = "drop")
  row_anno_f <- data.frame(
    row.names   = rownames(mat_f_norm),
    Total_Cells = total_cells_f$total_cells[
      match(rownames(mat_f_norm), total_cells_f$cell_type)
    ]
  )
  row_anno_f$Total_Cells[is.na(row_anno_f$Total_Cells)] <- 0
}

anno_colors_f <- list(
  Total_Cells = colorRampPalette(c("#deebf7","#08519c"))(100)
)

fig4f <- pheatmap(
  mat_f_norm,
  color            = viridis(100),
  cluster_rows     = FALSE,
  cluster_cols     = TRUE,
  show_rownames    = TRUE,
  show_colnames    = TRUE,
  fontsize_row     = 8,
  fontsize_col     = 7,
  border_color     = NA,
  main             = "NDD gene × cell type (0–1 normalised DEG count)",
  annotation_row   = row_anno_f,
  annotation_colors = anno_colors_f,
  silent           = TRUE
)

grid.newpage()
grid.draw(fig4f$gtable)

# pdf(file.path(OUT_DIR, "Fig4F_Heatmap_2k_perturb.pdf"), width = 11, height = 8)
# grid.newpage(); grid.draw(fig4f$gtable)
# dev.off()
```

---

## Figure 4H – Fisher enrichment: DEG sets × NDD gene sets

```{r fig4h, fig.width=11, fig.height=5}
# ── Load pre-computed NDD list ────────────────────────────────────────────────
ref_rds <- file.path(DATA_DIR, "ndd_perturbation", "ref_ndd_list_edit.RDS")
stopifnot(file.exists(ref_rds))
ref_ndd_list_edit <- readRDS(ref_rds)

# ── Build per-cell-type DEG lists ─────────────────────────────────────────────
de_res_lfc <- de_res[abs(de_res$logfoldchanges) > 0.5, ]

de_res_list <- de_res_lfc %>%
  group_by(group_name) %>%
  summarise(genes = list(unique(names)), .groups = "drop") %>%
  deframe()

# Background universe
univ_path <- file.path(DATA_DIR, "ndd_perturbation", "unique_genes_all.csv")
alt_univ  <- file.path(DATA_DIR, "ndd_perturbation", "all_genes.csv")
if (file.exists(alt_univ)) {
  bg_genes <- read.csv(alt_univ)$X
} else {
  stopifnot(file.exists(univ_path))
  bg_genes <- read.csv(univ_path)$names
}
de_res_list[["g_bg"]] <- bg_genes

desired_order_h <- c(CELLTYPE_ORDER, "g_bg")
desired_order_h <- intersect(desired_order_h, names(de_res_list))

# ── Fisher exact test helper ──────────────────────────────────────────────────
run_gene_enrich <- function(genesetA, genesetB) {
  bg_A <- sort(unique(unlist(genesetA)))
  bg_B <- sort(unique(unlist(genesetB)))
  rows <- vector("list", length(genesetA) * length(genesetB))
  idx  <- 1L
  for (nB in names(genesetB)) {
    gB    <- genesetB[[nB]]
    gB_bg <- bg_B[!bg_B %in% gB]
    for (nA in names(genesetA)) {
      gA    <- genesetA[[nA]]
      gA_bg <- bg_A[!bg_A %in% gA]
      a <- sum(gB %in% gA)
      b <- sum(gA_bg %in% gB)
      c <- sum(gB_bg %in% gA)
      d <- sum(gA_bg %in% gB_bg)
      ft <- fisher.test(matrix(c(a, b, c, d), ncol = 2), alternative = "greater")
      rows[[idx]] <- data.frame(
        setA = nA, setB = nB,
        setA_size = length(gA), setB_size = length(gB),
        setA_size_background = length(bg_A), setB_size_background = length(bg_B),
        setA_size_hits = a,
        fisher_OR = as.numeric(ft$estimate),
        fisher_p  = ft$p.value,
        overlaps  = paste(sort(intersect(gB, gA)), collapse = ", "),
        stringsAsFactors = FALSE
      )
      idx <- idx + 1L
    }
  }
  do.call(rbind, rows)
}

# ── Plot heatmap of enrichments ───────────────────────────────────────────────
plot_enrich_heatmap <- function(res_df, orderA, orderB, title = "") {
  df <- res_df %>%
    filter(setA != "g_bg", setB != "g_bg") %>%
    mutate(fisher_padj = p.adjust(fisher_p, method = "bonferroni", n = n()),
           setA_names = factor(setA, levels = orderA),
           setB_names = factor(setB, levels = rev(orderB)),
           OR_capped  = pmax(1/8, pmin(8, fisher_OR)))

  ggplot(df, aes(setA_names, setB_names)) +
    geom_tile(fill = "white", linewidth = 0.1, color = NA) +
    geom_point(aes(fill = log2(OR_capped)), shape = 22, size = 3,
               color = "white", stroke = 0.1) +
    geom_point(data = df[df$fisher_padj > 0.05 & df$fisher_p <= 0.05, ],
               aes(fill = log2(OR_capped)), shape = 22, size = 4,
               color = "white", stroke = 0.1) +
    geom_tile(data = df[df$fisher_padj <= 0.05, ],
              aes(fill = log2(OR_capped)), linewidth = 0.1, color = NA) +
    geom_point(data = df[df$fisher_padj <= 0.05, ],
               shape = 5, size = 1, color = "black") +
    scale_fill_gradient2(low = "#7FA6D9", mid = "white", high = "#E57373",
                         limits = c(log2(1/8), log2(8))) +
    labs(title = title, x = "", y = "") +
    coord_equal() +
    theme_minimal(base_size = 10) +
    theme(
      axis.ticks        = element_blank(),
      panel.border      = element_rect(linewidth = 1.5, fill = NA),
      panel.grid.major  = element_blank(),
      panel.grid.minor  = element_blank(),
      panel.background  = element_blank(),
      axis.text.x       = element_text(size = 10, angle = 320,
                                       hjust = 0, colour = "black"),
      axis.text.y       = element_text(size = 10, colour = "black"),
      legend.key.size   = unit(0.4, "cm")
    )
}

message("Running Fisher enrichment tests (may take a minute)...")
enrich_res <- run_gene_enrich(
  genesetA = de_res_list[desired_order_h],
  genesetB = ref_ndd_list_edit
)

fig4h <- plot_enrich_heatmap(
  enrich_res,
  orderA = desired_order_h,
  orderB = names(ref_ndd_list_edit),
  title  = "DEG set × NDD gene enrichment (Fisher exact, Bonferroni)"
)

fig4h

# ggsave(fig4h, filename = file.path(OUT_DIR, "Fig4H_FET_DEG_RiskGene.pdf"),
#        width = 11, height = 5)
```

---

## Figure 4I – NDD gene enrichment volcano per perturbation target

```{r fig4i, fig.width=9, fig.height=6}
gws_path  <- file.path(DATA_DIR, "ndd_perturbation", "GWS_results.csv")
asd_path  <- file.path(DATA_DIR, "ndd_perturbation", "ASD_rare.csv")
univ_path <- file.path(DATA_DIR, "ndd_perturbation", "unique_genes_all.csv")
stopifnot(file.exists(gws_path), file.exists(asd_path), file.exists(univ_path))

min_degs <- 20L

ndd_tbl  <- fread(asd_path, data.table = FALSE)
ndd_tbl  <- ndd_tbl[ndd_tbl$NDD373 == TRUE, ]
colnames(ndd_tbl)[1] <- "gene_symbol"
ndd_genes <- toupper(unique(ndd_tbl$gene_symbol))

univ_df  <- fread(univ_path, data.table = FALSE)
univ     <- sort(unique(toupper(univ_df$names)))
U        <- length(univ)

degs_gws <- fread(gws_path, data.table = FALSE)
degs_gws$gene_target <- toupper(as.character(degs_gws$gene_target))
degs_gws$gene_deg    <- toupper(as.character(degs_gws$names))

ndd_in_univ <- intersect(ndd_genes, univ)
n_ndd       <- length(ndd_in_univ)
message(sprintf("NDD genes in universe: %d / %d", n_ndd, length(ndd_genes)))

targets <- sort(unique(degs_gws$gene_target))

res_list <- lapply(targets, function(t) {
  tgenes <- setdiff(intersect(toupper(unique(degs_gws$gene_deg[degs_gws$gene_target == t])),
                              univ), t)
  n_t <- length(tgenes)
  if (n_t < min_degs) return(NULL)
  a <- sum(tgenes %in% ndd_in_univ)
  b <- n_t - a
  c <- n_ndd - a
  d <- U - a - b - c
  ft <- fisher.test(matrix(c(a, b, c, d), nrow = 2, byrow = TRUE))
  exp_a <- n_t * (n_ndd / U)
  data.frame(
    target          = t,
    n_target_genes  = n_t,
    overlap         = a,
    expected_overlap = exp_a,
    fold_enrichment = ifelse(exp_a > 0, a / exp_a, NA_real_),
    odds_ratio      = as.numeric(ft$estimate),
    p_value         = ft$p.value,
    pct_of_target   = 100 * a / n_t,
    stringsAsFactors = FALSE
  )
})

res_i <- do.call(rbind, Filter(Negate(is.null), res_list)) %>%
  arrange(p_value) %>%
  mutate(
    p_adj_bh      = p.adjust(p_value, method = "BH"),
    log2FE        = log2(fold_enrichment),
    negLog10_padj = -log10(p_adj_bh + 1e-300)
  )

# Save results
fwrite(res_i, file.path(OUT_DIR, "Fig4I_NDD_enrichment_results.csv"))

top_p  <- head(res_i %>% arrange(p_adj_bh), 10)$target
top_fe <- head(res_i %>% arrange(desc(fold_enrichment)), 10)$target
labels <- unique(c(top_p, top_fe))

fig4i <- ggplot(res_i, aes(x = log2FE, y = negLog10_padj)) +
  geom_point(aes(size = n_target_genes, color = pct_of_target), alpha = 0.85) +
  geom_text_repel(data = res_i %>% filter(target %in% labels),
                  aes(label = target), size = 3, max.overlaps = 50) +
  scale_size_continuous(name = "n DEGs (target)") +
  scale_color_viridis(name = "% target DEGs in NDD", option = "viridis") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  labs(
    x        = "log2(Fold enrichment vs expectation)",
    y        = "-log10(BH-adjusted p-value)",
    title    = "Overlap of perturbation target DEG sets with NDD signature",
    subtitle = sprintf("min_degs = %d | Universe = %d | NDD_in_univ = %d",
                       min_degs, U, n_ndd)
  ) +
  theme_classic(base_size = 12) +
  theme(panel.grid = element_blank())

fig4i

# ggsave(fig4i, filename = file.path(OUT_DIR, "Fig4I_DEG_NDD_Enrichment.pdf"),
#        width = 9, height = 6, dpi = 300)
```
