---
title: "Figure 3 — Grin/Gria Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 14,
  fig.height = 5.5
)
```

```{r libraries}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(readxl)
  library(arrow)
  library(pheatmap)
  library(RColorBrewer)
  library(scales)
})
```

```{r paths}
# All data relative to this Rmd file
DATA_DIR <- "/data/wholebrain_crispr_atlas"

PARQUET <- file.path(DATA_DIR, "analysis", "diff_expr_genes", "wilcoxon_de_results_group_name_final_data_no_multi_guide_cells.parquet")
```

```{r read-parquet}
df_raw <- read_parquet(PARQUET)
```

---

### Figure 3D — Grin/Gria Dotplot: effect size across brain regions

```{r fig3d-data}
grin_de <- df_raw %>%
  filter(pvals_adj < 0.1) %>%
  filter(grepl("^grin|^gria", gene_target, ignore.case = TRUE)) %>%
  mutate(group = gsub("-", " ", group_name, fixed = TRUE))

summary_df <- grin_de %>%
  group_by(gene_target, group) %>%
  summarize(
    n_deg        = n(),
    mean_abs_lfc = mean(abs(logfoldchanges), na.rm = TRUE),
    .groups = "drop"
  )

grin_levels  <- sort(unique(summary_df$gene_target))
group_levels <- sort(unique(summary_df$group))

plot_df <- tidyr::expand_grid(
  gene_target = grin_levels,
  group       = group_levels
) %>%
  left_join(summary_df, by = c("gene_target", "group")) %>%
  filter(!is.na(n_deg))

cat("Unique gene targets:", paste(sort(unique(plot_df$gene_target)), collapse = ", "), "\n")
cat("Unique cell-type groups:", length(unique(plot_df$group)), "\n")
```

```{r fig3d-plot, fig.width=16, fig.height=5}
row_order <- rev(c("Grin1", "Grin2a", "Grin2b", "Grin2d", "Gria2", "Gria1", "Gria3"))
col_order <- c(
  "007 L2 3 IT CTX Glut", "005 L4 5 IT CTX Glut", "022 L5 ET CTX Glut",
  "151 TH Prkcd Grin2c Glut", "155 MB Glut", "052 Pvalb Gaba",
  "145 MH LH TH Glut", "053 Sst Gaba", "008 L2 3 IT ENT PPP RSP Glut",
  "191 MB P MY GABA", "217 P MY Pineal Glut", "001 L5 6 IT Glut",
  "110 CNU HYa HY MM Glut", "066 CNU HYa HY GABA", "046 CTX CGE GABA",
  "027 NP CT L6b OB Glut", "009 L2 3 IT PIR AON ENT Glut", "308 CB GABA"
)

# Runtime checks — warn instead of silently dropping
missing_rows <- setdiff(row_order, unique(plot_df$gene_target))
missing_cols <- setdiff(col_order, unique(plot_df$group))
if (length(missing_rows) > 0) warning("row_order entries not in data: ", paste(missing_rows, collapse = ", "))
if (length(missing_cols) > 0) warning("col_order entries not in data: ", paste(missing_cols, collapse = ", "))

deg_breaks  <- c(10, 50, 100)
size_breaks <- log10(deg_breaks + 1)
size_labels <- as.character(deg_breaks)

plot_df_f <- plot_df %>%
  mutate(
    n_deg_log10 = log10(n_deg + 1),
    gene_target = factor(gene_target, levels = row_order),
    group       = factor(group,       levels = col_order)
  )

p3d <- ggplot(plot_df_f, aes(x = group, y = gene_target)) +
  geom_tile(fill = NA, color = NA) +
  geom_point(
    aes(size = n_deg_log10, fill = mean_abs_lfc),
    shape  = 21,
    color  = "grey40",
    stroke = 0.6,
    alpha  = 0.95,
    na.rm  = TRUE
  ) +
  scale_size_continuous(
    name   = "#DEGs",
    range  = c(3, 10),
    breaks = size_breaks,
    labels = size_labels
  ) +
  scale_fill_gradientn(
    name   = "avg(abs(log2FC))",
    colors = c("white", "#6baed6", "#2171b5", "#08519c"),
    values = rescale(c(0, 0.4, 0.7, 1.0))
  ) +
  labs(
    x     = NULL,
    y     = NULL,
    title = "Cell type-restricted effects of NMDA receptor signaling"
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x  = element_text(angle = 60, hjust = 1, vjust = 1, size = 8.5),
    axis.text.y  = element_text(face = "italic", size = 12),
    axis.ticks   = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
    legend.position  = "right",
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 9),
    plot.title   = element_text(face = "bold", size = 13, hjust = 0)
  ) +
  guides(
    size = guide_legend(order  = 1),
    fill = guide_colorbar(order = 2)
  )

p3d
```

---

## Figure 3E — Grin2a / Grin2b Correlation Plot (L2-3 IT CTX Glut)

Uses parquet DEG data (`df_raw`), subset to gene_target Grin2a and Grin2b.

```{r fig3e-data}
grin_de_sub <- df_raw %>%
  filter(grepl("^grin", gene_target, ignore.case = TRUE)) %>%
  mutate(group = gsub("-", " ", group_name, fixed = TRUE))

cat("Columns:", paste(names(grin_de_sub), collapse = ", "), "\n")
cat("Unique groups:", length(unique(grin_de_sub$group)), "\n")
```

```{r fig3e-plot, fig.width=16, fig.height=5}
target_group <- "007 L2 3 IT CTX Glut"

sub <- grin_de_sub %>%
  filter(group == target_group, gene_target %in% c("Grin2a", "Grin2b")) %>%
  group_by(gene_target, names) %>%
  summarize(
    lfc  = mean(logfoldchanges, na.rm = TRUE),
    padj = min(pvals_adj,       na.rm = TRUE),
    .groups = "drop"
  )

grin2a  <- sub %>% filter(gene_target == "Grin2a") %>%
  transmute(names, lfc_grin2a = lfc, padj_grin2a = padj)
grin2b  <- sub %>% filter(gene_target == "Grin2b") %>%
  transmute(names, lfc_grin2b = lfc, padj_grin2b = padj)

overlap <- inner_join(grin2a, grin2b, by = "names") %>%
  mutate(
    direction = ifelse(lfc_grin2a * lfc_grin2b >= 0, "same", "opposite"),
    sig_a     = padj_grin2a <= 0.1,
    sig_b     = padj_grin2b <= 0.1,
    sig_class = case_when(
      sig_a & sig_b   ~ "both",
      sig_a & !sig_b  ~ "Grin2a only",
      !sig_a & sig_b  ~ "Grin2b only",
      TRUE            ~ "neither"
    ),
    evidence  = -log10(pmin(padj_grin2a, padj_grin2b))
  ) %>%
  filter(sig_a | sig_b)

cat("Shared significant genes:", nrow(overlap), "\n")

cor_res <- if (nrow(overlap) >= 2 && sum(complete.cases(overlap[, c("lfc_grin2a", "lfc_grin2b")])) >= 2) {
  cor.test(overlap$lfc_grin2a, overlap$lfc_grin2b, method = "pearson")
} else {
  list(estimate = NA_real_, p.value = NA_real_)
}

n_labeled_approx <- min(nrow(overlap), 30)

p3e <- ggplot(overlap, aes(x = lfc_grin2a, y = lfc_grin2b)) +
  geom_hline(yintercept = 0, linewidth = 0.35, color = "#9aa1a8") +
  geom_vline(xintercept = 0, linewidth = 0.35, color = "#9aa1a8") +
  geom_point(aes(color = direction), size = 3.8, alpha = 0.95) +
  geom_smooth(
    method   = "lm", se = TRUE,
    color    = "black", fill = "grey80",
    linetype = "dashed", linewidth = 0.6
  ) +
  annotate(
    "text",
    x     = -Inf, y = Inf,
    label = paste0(
      "Pearson r = ", round(cor_res$estimate, 2),
      "\np = ", signif(cor_res$p.value, 3)
    ),
    hjust  = -0.1,
    vjust  = 1.3,
    size   = 4.5
  ) +
  geom_text(
    aes(label = names),
    nudge_x      = 0.03,
    nudge_y      = 0.03,
    size         = 4.2,
    check_overlap = TRUE
  ) +
  scale_color_manual(values = c("opposite" = "#1f77b4", "same" = "#d62728")) +
  labs(
    title = paste0(target_group, ": Grin2a vs Grin2b shared DEGs"),
    x     = "Grin2a log2FC",
    y     = "Grin2b log2FC",
    color = NULL
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold"),
    legend.position  = "top"
  )

p3e
```

---

## Data File 3 — `wilcoxon_de_results_group_name_final_data_no_multi_guide_cells.parquet`

### Figure 3F — Shared DEG Heatmap across perturbations (L2-3 IT CTX Glut)

**Notes:**
- `values_fill = NA` instead of `0` to avoid imputing zero for untested pairs.
- Z-scores alone are used for the heatmap.

```{r fig3f-data}
sample_groups <- head(unique(df_raw$group_name), 5)
cat("Sample group_name values:", paste(sample_groups, collapse = " | "), "\n")

TARGET_GROUP <- "007 L2-3 IT CTX Glut"
PERTS_F      <- c("Grin2a", "Gabrg2", "Kcnb1", "Grin2b", "Ap2s1", "Ap2m1")

if (!TARGET_GROUP %in% unique(df_raw$group_name)) {
  stop("group_name '", TARGET_GROUP, "' not found. Check hyphen vs space format.")
}

df <- df_raw %>%
  filter(group_name == TARGET_GROUP, gene_target %in% PERTS_F)

df_degs <- unique(df[df$gene_target %in% c("Grin2a", "Grin2b") & df$pvals_adj < 0.05, ]$names)
cat("DEG universe size (padj < 0.05, Grin2a|Grin2b):", length(df_degs), "\n")

df_sub <- df %>%
  filter(names %in% df_degs) %>%
  select(names, logfoldchanges, gene_target)
```

```{r fig3f-plot, fig.width=16, fig.height=5}
mat <- df_sub %>%
  pivot_wider(
    names_from   = gene_target,
    values_from  = logfoldchanges,
    values_fill  = NA
  ) %>%
  column_to_rownames("names") %>%
  as.matrix()

col_keep <- intersect(PERTS_F, colnames(mat))
if (length(setdiff(PERTS_F, colnames(mat))) > 0)
  warning("Perturbations missing from matrix: ",
          paste(setdiff(PERTS_F, colnames(mat)), collapse = ", "))
mat <- mat[, col_keep, drop = FALSE]

mat <- mat[rowSums(!is.na(mat)) > 0, , drop = FALSE]

mat_z <- t(scale(t(mat)))

bad_rows <- apply(mat_z, 1, function(x) all(is.nan(x) | is.na(x)))
if (any(bad_rows)) {
  message(sum(bad_rows), " constant/all-NA gene rows removed before clustering.")
  mat_z <- mat_z[!bad_rows, , drop = FALSE]
}

heat_colors <- colorRampPalette(c("#3b6fb6", "white", "#c43c39"))(100)

pheatmap(
  mat_z,
  color                    = heat_colors,
  cluster_rows             = TRUE,
  cluster_cols             = FALSE,
  clustering_distance_rows = "correlation",
  scale                    = "none",
  border_color             = NA,
  show_rownames            = TRUE,
  show_colnames            = TRUE,
  fontsize_row             = 9,
  fontsize_col             = 13,
  angle_col                = 45,
  treeheight_row           = 60,
  main                     = paste0(TARGET_GROUP, ": shared Grin2a/Grin2b DEG heatmap\n(z-scored LFC, padj < 0.05)")
)
```
